---
title: "Cyclistic bikeshare analytics"
author: "James Paul Pulayan"
date: "2023-01-19"
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```
## **Ask**

This phase of the data process is about identifying the problems which are relevant to the stakeholders. This problems shall be the guide in carrying out each of the steps of the data processing.

### Background

This case study is one of the optional projects of track 1 of the "Google Data Analytics Capstone: Complete a Case Study". The course is the last part of the Google Data Analytics Professional Certification. This case study was described to be similar to case studies conducted on job interviews. The business task, data set and collections of identified deliverable are specified by the course and must be presented to expected stakeholders. Once done with the case study, it can be included in the portfolio to demonstrate data analytics related skills for future job interviews.
  
In the process of doing this case study, real-world tasks of a junior data analyst shall be completed. In answering key business questions for the devised company Cyclistic, the data analysis process: ask, prepare, process, analyze, share and act shall be followed.

### Scenario

I am part of the marketing analyst team of Cyclistic, a bike-share company in Chicago, as a junior data analyst. The director, Lily Moreno, believes that maximizing the number of annual memberships is relevant in the eventual success of the company. The finance analysts inferred that annual members are much more profitable than casual riders.
  
The bike-share program of Cyclistic features more than 5800 bicycles and 600 dock stations as well as assistive bikes like reclining bikes, hand tricycles and cargo bikes. About 8% of the riders utilizes the assistive bikes. Majority of the bikers use bikes for recreation and fun and about 30% use bikes for commuting.
  
The Cyclistic executive team shall determine the suitability of the recommended marketing program.

### Business use case

Guide questions for the business task:

How do annual members and casual riders use Cyclistic bikes differently? (The question asked by Moreno)

####  *Business task*
  * Formulate recommendations on how casual riders may upgrade to annual membership by understanding their differences from the result of the analysis of the historical data.

## **Data preparation**

The data source provides monthly data of users with the latest data dating to November of 2022.

### Setting up environment

```{r environment}
library(tidyverse) #For cleaning data
library(dplyr) #Data manipulation
library(skimr) #Data summary
library(lubridate) #Date manipulation
library(ggpattern) #for fill pattern
library(RColorBrewer) #Color palettes
library(rmarkdown) #documentation
```

Data from the previous 12 months of Cyclistic trip data shall be used. The datasets that will be included in the data analysis process are the trip data from January 1, 2021 up to November 30, 2022, 12 months in total. The data is imported to Rstudio:

```{r data import, message=FALSE}
tripdata_202112 <- read_csv("202112-divvy-tripdata.csv") #Dec, 2021
tripdata_202201 <- read_csv("202201-divvy-tripdata.csv") #Jan, 2022
tripdata_202202 <- read_csv("202202-divvy-tripdata.csv") #Feb, 2022
tripdata_202203 <- read_csv("202203-divvy-tripdata.csv") #Mar, 2022
tripdata_202204 <- read_csv("202204-divvy-tripdata.csv") #Apr, 2022
tripdata_202205 <- read_csv("202205-divvy-tripdata.csv") #May, 2022
tripdata_202206 <- read_csv("202206-divvy-tripdata.csv") #Jun, 2022
tripdata_202207 <- read_csv("202207-divvy-tripdata.csv") #Jul, 2022
tripdata_202208 <- read_csv("202208-divvy-tripdata.csv") #Aug, 2022
tripdata_202209 <- read_csv("202209-divvy-publictripdata.csv") #Sep, 2022
tripdata_202210 <- read_csv("202210-divvy-tripdata.csv") #Oct, 2022
tripdata_202211 <- read_csv("202211-divvy-tripdata.csv") #Nov, 2022
```

The data types are compared and checked. Each have similar structures:
* ride_id as character
* rideable_type as character
* started_at as datetime
* ended_at as datetime
* start_station_name as character
* start_station_id as character
* end_station_name as character
* end_station_id as character
* start_lat as numeric
* start_lng as numeric
* end_lat as numeric
* end_lng as numeric
* member_casual as character

To explore and analyze the data as a whole, the 12 data sets should be joined.

```{r message=FALSE, warning=FALSE}
tripdata_all <- tripdata_202112 %>% 
  full_join(tripdata_202201) %>% 
  full_join(tripdata_202202) %>% 
  full_join(tripdata_202203) %>% 
  full_join(tripdata_202204) %>% 
  full_join(tripdata_202205) %>% 
  full_join(tripdata_202206) %>% 
  full_join(tripdata_202207) %>%
  full_join(tripdata_202208) %>%
  full_join(tripdata_202209) %>%
  full_join(tripdata_202210) %>%
  full_join(tripdata_202211)
```

This results in a data frame, `tripdata_all`, with 5733451 rows, excluding the header column, and 13 columns. Then, the csv file is exported in order to have a copy of the raw data in the computer.

```{r trip data export, eval=FALSE, include=FALSE}
tripdata_all <- read_csv("tripdata_all.csv")
```


```{r}
nrow(tripdata_all) == n_unique(tripdata_all$ride_id)
```

The number of rows of `tripdata_all` is also equal to the number of unique values of `tripdata_all$ride_id` hence, it is the primary key.

## **Data processing**

The summary statistics is shown below:

```{r echo=FALSE}
skim_without_charts(tripdata_all)
```

### Conversion of categorical columns as factors

The `rideable_type` has 3 unique values namely as `electric_bike`, `classic_bike` and `docked_bike`. The `member_casual` has 2 unique values `member`, for users with annual subscription, and `casual`, for users who avail single-ride pass or full-day pass. The variables `rideable_type` and `member_casual` are converted as factors for better analysis using `ggplot2` using these variables.

```{r}
tripdata_all2 <- tripdata_all #Creating an alternative data frame for the conversion of data types
tripdata_all2$rideable_type <- as.factor(tripdata_all2$rideable_type)
tripdata_all2$member_casual <- as.factor(tripdata_all2$member_casual)
```


```{r include=FALSE}
#Checking for levels of factors
levels(tripdata_all2$rideable_type)
levels(tripdata_all2$member_casual)
```

### Cleaning of rows with missing values
Rows were checked for missing values of station names and station ids.

```{r}
#Checking for rows with `NA` values
##start station name has a value and start station id is missing
tripdata_all2 %>% 
  filter(!is.na(start_station_name)&is.na(start_station_id))

##start station name is missing and start station id has a value
tripdata_all2 %>% 
  filter(is.na(start_station_name)&!is.na(start_station_id))

##end station name has a value and end station id is missing
tripdata_all2 %>% 
  filter(!is.na(end_station_name)&is.na(end_station_id))

##end station name is missing and end station id has a value
tripdata_all2 %>% 
  filter(is.na(end_station_name)&!is.na(end_station_id))

```

There are no rows with a value on station name and a missing value on the station id and there are no rows with a value on station id and a missing value on the station name hence, missing values on station names and station ids appear simultaneously.

```{r}
tripdata_all2 %>% 
  filter(is.na(start_station_name)&is.na(start_station_id)) %>% 
  count(rideable_type, sort = TRUE)
tripdata_all2 %>% 
  filter(is.na(end_station_name)&is.na(end_station_id)) %>% 
  count(rideable_type, sort = TRUE)
```

All rows with missing values on both `start_station_name` and `start_station_id` have a `rideable_type` of `electric_bike` and a big percentage of the rows with missing values on both `end_station_name` and `end_station_id` have a `rideable_type` of `electric_bike`. This means that the error resulting in these missing values mostly occurs for electric bike users hence, these rows cannot be immediately deleted as deleting these rows may skew the results.

```{r}
tripdata_all2 %>% 
  filter(is.na(end_lat)&!is.na(end_lng))
tripdata_all2 %>% 
  filter(!is.na(end_lat)&is.na(end_lng))

```

Missing values on end latitude and longitude appear simultaneously as well. 
```{r}
tripdata_all2 %>% 
  filter(is.na(end_lat)&is.na(end_lng)) %>% 
  filter(!is.na(end_station_name))
```

These missing rows with missing values have missing values on the end station names and ids as well.

```{r}
trip_clean1 <- tripdata_all2 %>% 
  filter(!is.na(end_lat)&!is.na(end_lng))
```

Rows with missing values on latitude and longitude were deleted as the coordinates in the dataset are specific with the position of the user and not the station. There are 5874 rows deleted. The summary of the cleaned data set is shown:

```{r echo=FALSE}
skim_without_charts(trip_clean1)
```

### Checking of station names and ids

The unique start station names and start station ids were examined. A data frame showing the count of each unique ids of the station names sis shown below:

```{r}
dataframe1 <- trip_clean1 %>% 
  select(start_station_name, start_station_id) %>%
  group_by(start_station_name) %>% 
  summarise(ss_id_count = n_unique(start_station_id)) %>% 
  arrange(desc(ss_id_count))

dataframe1.1 <- trip_clean1 %>% 
  group_by(start_station_name) %>% 
  filter(n_unique(start_station_id)==2) %>% 
  summarise(start_station_id = unique(start_station_id))

dataframe1.2 <- trip_clean1 %>% 
  select(start_station_name, start_station_id) %>% 
  count(start_station_name, start_station_id)
```
```{r}
dataframe1.3 <- inner_join(dataframe1.1, dataframe1.2) %>% 
  arrange(desc(n), .by_group = TRUE)
```

There are 18 station names with two station ids and the difference on the count of station ids for each unique station names is large.

```{r}
dataframe2 <- trip_clean1 %>% 
  select(start_station_id, start_station_name) %>%
  group_by(start_station_id) %>% 
  summarise(ss_name_count = n_unique(start_station_name)) %>% 
  arrange(desc(ss_name_count))

dataframe2.1 <- trip_clean1 %>% 
  group_by(start_station_id) %>% 
  arrange(desc(n_unique(start_station_name))) %>% 
  filter(n_unique(start_station_name)!=1) %>%
  summarise(start_station_name = unique(start_station_name))

dataframe2.2 <- trip_clean1 %>% 
  select(start_station_id, start_station_name) %>% 
  count(start_station_id, start_station_name)

dataframe2.3 <- inner_join(dataframe2.1, dataframe2.2) %>% 
  arrange(desc(n), .by_group = TRUE)
```

The greatest count of unique station names is 4 which is the case for the station id `564`.

```{r echo=FALSE}
dataframe2.3[209:212,]
```

There are also station ids with counts of unique stations names of less than 4.

Some start station ids with multiple station names contains station names that are similar, e.g.:

```{r echo=FALSE}
dataframe2.3[41:50,]
```

Some start station ids have misspellings on their station names resulting in multiple counts of unique start station names.

```{r echo=FALSE}
dataframe2.3[3:6,]
```

It is decided to use station ids as the identification for the stations. The table below shows the new start station ids for each start station names with two unique start station ids. The start station ids with the higher count was used. The new ids for each start station names with two station ids is shown below:

```{r}
dataframe1.4 <- data.frame(dataframe1.3[seq(1,36,by=2),1:2])

dataframe1.5 <- trip_clean1 %>% 
  group_by(start_station_name) %>% 
  filter(n_unique(start_station_id)==1) %>% 
  summarise(start_station_id = unique(start_station_id))

dataframe1.6 <- full_join(dataframe1.4, dataframe1.5)
start_station_names_ids <- dataframe1.6
```

The data set is cleaned with the new start station ids. The first 6 rows of the new data set is shown: 

```{r cleaning start_station_name_id}
trip_clean2 <- left_join(trip_clean1, start_station_names_ids, by = "start_station_name", suffix = c(".x",""))
trip_clean2 <- trip_clean2 %>% 
  select(1:5,14,7:13)
head(trip_clean2)
```

Next, the same data cleaning process is applied with the end station names and ids. The table of the end station names with two unique ids and their count is shown:

```{r}
dataframe3 <- trip_clean2 %>% 
  select(end_station_name, end_station_id) %>%
  group_by(end_station_name) %>% 
  summarise(es_id_count = n_unique(end_station_id)) %>% 
  arrange(desc(es_id_count))

dataframe3.1 <- trip_clean2 %>% 
  group_by(end_station_name) %>% 
  filter(n_unique(end_station_id)==2) %>% 
  summarise(end_station_id = unique(end_station_id))

dataframe3.2 <- trip_clean2 %>% 
  select(end_station_name, end_station_id) %>% 
  count(end_station_name, end_station_id)

dataframe3.3 <- inner_join(dataframe3.1, dataframe3.2) %>% 
  arrange(desc(n), .by_group = TRUE)
dataframe3.3
```

The table below shows the new end station ids for each end station names with two unique end station ids. The end station ids with the higher count was used similar with the first process. The new ids for each end station names with two station ids is shown below:

```{r}
dataframe3.4 <- data.frame(dataframe3.3[seq(1,36,by=2),1:2])

dataframe3.5 <- trip_clean2 %>% 
  group_by(end_station_name) %>% 
  filter(n_unique(end_station_id)==1) %>% 
  summarise(end_station_id = unique(end_station_id))

dataframe3.6 <- full_join(dataframe3.4, dataframe3.5)
end_station_names_ids <- dataframe3.6
```

The data set is cleaned with the new end station ids. The first 6 rows of the new data set is shown: 

```{r cleaning end_station_name_id}
trip_clean3 <- left_join(trip_clean2, end_station_names_ids, by = "end_station_name", suffix = c(".x",""))
trip_clean3 <- trip_clean3 %>% 
  select(1:7,14,9:13)
head(trip_clean3)
```

The columns containing station ids are converted to factors. The station  names columns are removed, the station ids columns shall be used in grouping by location.

```{r}
##Conversion of station id columns to factors
trip_clean3$start_station_id  <- as_factor(trip_clean3$start_station_id)
trip_clean3$end_station_id  <- as_factor(trip_clean3$end_station_id)

##Removal of the station name columns
trip_clean4 <- trip_clean3 %>% 
  select(1:4,6,8:13)
head(trip_clean4)
```

### Cleaning of the grid coordinate columns

A scatter plot is made with the `start_lat` as the horizontal values and `start_lng` as the vertical values.

```{r}
scatterplot1 <- ggplot(trip_clean4, aes(x=start_lat, y=start_lng))+
  geom_jitter(alpha = 0.4)
scatterplot1
```

The points are cluttered between $41.5^{\circ}$ and $42.5^{\circ}$ latitude and an outlier is at `start_lat>45`.

Another scatter plot is made with the `end_lat` as the horizontal values and `end_lng` as the vertical values. The same with the first scatterplot, outliers are identified where `end_lat` and `end_lng` are both zero.

```{r}
scatterplot2 <- ggplot(trip_clean4, aes(x=end_lat, y=end_lng))+
  geom_jitter(alpha = 0.4)
scatterplot2
```

The [border coordinates](https://en.wikipedia.org/wiki/Module:Location_map/data/United_States_Chicago) of Chicago for latitudes are $41.9653^{\circ}$(top) and $41.8096^{\circ}$(bottom) and for longitudes $-87.7738^{\circ}$(left) and $-87.5658^{\circ}$(right). These outliers are removed, 9 rows are deleted.

```{r}
trip_clean5 <- trip_clean4 %>% 
  filter(end_lat!=0) %>% 
  filter(between(start_lat, 41.5, 42.5))
head(trip_clean5)
```

### Addition of two new variables day and month

The initial time of ride, represented by the variable `started_at`, and the final time of ride, represented by the variable `ended_at`, are POSIXct date times. A new variable is added which measures the duration of a trip. It is the difference between `ended_at` and `started_at`. Negative duration were discarded, 10 rows are deleted.

```{r}
trip_new1 <- trip_clean4 %>% 
  mutate(ride_length = int_length(interval(started_at, ended_at)),
         .after = ended_at) %>% 
  filter(ride_length >= 0)
```

Two columns are added which are the day and month of the starting datetime of a ride. The new columns are ordered factor data types.

```{r}
trip_data <- trip_new1 %>% 
  mutate(day = wday(started_at, label = TRUE, abbr = FALSE),
         month = month(started_at, label = TRUE, abbr = FALSE))
head(trip_data)
```

After processing and cleaning the raw data, a summary for the cleaned data set is shown:

```{r}
skim_without_charts(trip_data)
```

